---
title: "Co-occurrence Graphs"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
```

## Co-occurrence graphs

We are going to look at microbial abundances of different strains/species of bacteria in biological samples.

These samples come from a longitudinal study of mice from several litters.


### Bacterial community example using a `phyloseq` object

We load the `phyloseq` package first and read the phyloseq object that integrates all the data.

```{r phyloseqnetwork0,echo=TRUE}
library("phyloseq")
#Read an external file
micephyloseq  = readRDS(url("https://web.stanford.edu/class/bios221/data/ps1.rds"))
#Internal file from the package is also available 
# data(micephyloseq)
micephyloseq
```

We could like to connect the samples that have a lot of bacteria in common.
The best way to measure similarity between species abundances is the "Jaccard distance".

We start by computing the Jaccard distance between samples and set a threshold for the distance to correspond to an edge in the "co-occurrence" graph.

Exercise: Add a line of code that makes the shape give the litter information:

```{r phyloseqnetwork1,echo=TRUE,exercise=TRUE,eval.exercise=TRUE}
data(micephyloseq)
micephyloseq
ig<-make_network(micephyloseq, max.dist=0.3)
plot_network(ig,micephyloseq,color="host_subject_id")
```


```{r phyloseqnetwork1-solution}
ig<-make_network(micephyloseq, max.dist=0.3)
plot_network(ig,micephyloseq,color="host_subject_id",shape="family_relationship")
```

We can replicate what the function is doing by hand,
using the tidygraph package, you have to add a few lines
to remove the isolates.

We start by loading the libraries
```{r loadl,message=FALSE}
library("tidygraph")
library("igraph")
library("ggraph")
```


```{r JaccardThreshold,exercise=TRUE}
d1 <- as.matrix(phyloseq::distance(micephyloseq, method="jaccard"))
d1 <- d1 +diag(ncol(d1))
threshM <- (d1<0.35)
gr <- graph_from_adjacency_matrix(threshM,  mode = "lower")
sampledata = data.frame( sample_data(micephyloseq))
V(gr)$id = sampledata[names(V(gr)), "host_subject_id"]
V(gr)$litter = sampledata[names(V(gr)), "family_relationship"]
tigr<- gr%>% as_tbl_graph()
tigr
ggraph(tigr,layout="fr")+ geom_edge_link() +
  geom_node_point(aes(color=id,shape=litter))+
  theme_graph()
```

```{r JaccardThreshold-solution}
tigr %>%
  activate(nodes) %>% 
  filter(!node_is_isolated()) %>%
  ggraph(,layout="fr") + geom_edge_link() +
  geom_node_point(aes(color=id,shape=litter))+
  theme_graph()
```
