---
title: "Using Graphs for Statistics"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE, message=FALSE)
```

## Testing a graph for association with a factor variable

For univariate data in two groups, we can test the differences by looking at the number of runs within each group.

Seeing the number of runs in a one-dimensional, two-sample, nonparametric Wald-Wolfowitz test can indicate whether the two groups have the same distributions.

```{r WWtest, echo=FALSE, fig.width=10, fig.height=2}
library(ggplot2)
dfbr=data.frame(measure=c(rnorm(15,0.9),rnorm(15,1.8)),
  group=as.factor(c(rep("men",15),rep("women",15))))
ggplot(dfbr,aes(x=measure,group=group,y=0)) + ylim(-0.25,+0.25) +
  geom_point(aes(col=group,x=measure,y=0,shape=group),size=5,alpha=0.6)+
  scale_color_manual(values=c("blue","red"))+
  theme_bw() + geom_hline(yintercept = 0) +
  theme(panel.border = element_blank(),
  axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        panel.grid.major.y = element_blank() ,
        panel.grid.minor.y = element_blank() )+ coord_fixed()
```




## Testing the relation between a graph and vertex factors

### The samples have factor covariates

The nodes in the graph correspond to
the samples and are associated to
factors that specify the litter and individual ids of the mice from which the samples were collected.

```{r phylograph}
library("phyloseq")
library("phyloseqGraphTest")
library("igraph")
data(micephyloseq)
#micephyloseq  = readRDS(url("https://web.stanford.edu/class/bios221/data/ps1.rds"))
micephyloseq
sampledata = data.frame( sample_data(micephyloseq))
d1 = as.matrix(phyloseq::distance(micephyloseq, method="jaccard"))
gr = graph.adjacency(d1,  mode = "undirected", weighted = TRUE)
net = igraph::mst(gr)
V(net)$id = sampledata[names(V(net)), "host_subject_id"]
V(net)$litter = sampledata[names(V(net)), "family_relationship"]
```


The minimum spanning tree based on Jaccard dissimilarity and annotated with the mice ID and litter factors.


```{r mstplot, fig.width=4, fig.height=4.5}
library(ggnetwork)
gnet=ggnetwork(net)
ggplot(gnet, aes(x = x, y = y, xend = xend, yend = yend))+
  geom_edges(color = "darkgray") +
  geom_nodes(aes(color = id, shape = litter)) + theme_blank()+
  theme(legend.position="bottom")
```

```{r MSTJaccardplain}
library("phyloseqGraphTest")
gt = graph_perm_test(micephyloseq, "host_subject_id", distance="jaccard",
                     type="mst",  nperm=1000)
plot_test_network(gt)
```

```{r testresults}
gt
plot_permutations(gt)
gt$pval

```
The permutation histogram of the number of pure edges in the network obtained from the minimal spanning tree with Jaccard similarity.


```{r ggnetworkphyl}
net = make_network(micephyloseq, max.dist = 0.35)
sampledata = data.frame(sample_data(micephyloseq))
V(net)$id = sampledata[names(V(net)), "host_subject_id"]
V(net)$litter = sampledata[names(V(net)), "family_relationship"]
netg = ggnetwork(net)
```


```{r ggnetworkplotJ,  fig.cap = "A co-occurrence network created by using a threshold on the Jaccard dissimilarity matrix. The colors represent which mouse the sample came from; the shape represents which litter the mouse was in.", fig.margin = FALSE, fig.width=6, fig.height=5}
ggplot(netg, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(color = "darkgray") +
  geom_nodes(aes(color = id, shape = litter)) + theme_blank()+
    theme(plot.margin = unit(c(0, 5, 2, 0), "cm"))+
    theme(legend.position = c(1.4, 0.3),legend.background = element_blank(),
          legend.margin=margin(0, 3, 0, 0, "cm"))+
         guides(color=guide_legend(ncol=2))
```

```{r mstperm}
gt = graph_perm_test(micephyloseq, "family_relationship",
        grouping = "host_subject_id",
        distance = "jaccard", type = "mst", nperm= 1000)
gt$pval
```


```{r mstpermplotNest, fig.keep = 'high', fig.cap = "The permutation histogram obtained from the minimal spanning tree with Jaccard similarity.", fig.width=4, fig.height=2.1}
plot_permutations(gt)
```

```{r knn1test}
gtnn1 = graph_perm_test(micephyloseq, "family_relationship",
                      grouping = "host_subject_id",
                      distance = "jaccard", type = "knn", knn = 1)
gtnn1$pval
```

```{r knn1plot, fig.keep = 'high', fig.cap = "The graph obtained from a nearest-neighbor graph with Jaccard similarity.", fig.width=4, fig.height = 2.7}
plot_test_network(gtnn1)

```




References:

 - [phyloseqGraphTest vignette](https://cran.r-project.org/web/packages/phyloseqGraphTest/vignettes/gt_vignette.html)
 
 - [Friedman Rafsky explanation and details](https://web.stanford.edu/class/bios221/book/Chap-Graphs.html#mst-based-testing-the-friedmanrafsky-test)
 
 