---
title: "Layout for Graphs"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
```


## Choices made by the plotting programs

### Different graph layouts

## Graph Layout

```{r igraphload}
library("igraph")
```

```{r graphprepare}
gr<-readRDS("../gr.RDS")
```

If you look at the help for `igraph.plotting` you find a description of many different choices for *laying out* the graph as usually the x and y  coordinates are not provided 
data but extra information that is only computed for aesthetic or clarity.

If you don't provide a layout all the graph plotting functions will pick one for your
(very often a Fruchterman-Reingold) which tries to stretch the graph enough so that are not too many overlapping edges, a little like Multidimensional Scaling stretches the points out.

Let's look at some examples; these are all the same graphs but with different layout choices.

```{r layout1, echo=TRUE}
co <- layout_with_fr(gr)
## Fruchterman-Reingold layout (force directed layout)
plot(gr, layout=co)
## On a circle
coordsc <- layout_in_circle(gr)
plot(gr, layout=coordsc)
##The Kamada-Kawai layout algorithm
plot(gr, layout=layout_with_kk)
```

Try to find other layout choices
(on a grid, on a sphere, through multidimensional scaling).




```{r layoutchoice, exercise=TRUE,exercise.lines = 6}

```


```{r layoutchoice-hint}
lgrid <- layout_on_grid(gr)
lmds <- layout_with_mds(gr)
lsphere <- layout_on_sphere(gr)
```

```{r layoutchoice-solution}
lmds <- layout_with_mds(gr)
lgrid <- layout_on_grid(gr)
lsphere <- layout_on_sphere(gr)
plot(gr, layout=lmds)
plot(gr,layout=lgrid)
plot(gr, layout=lsphere)
```


### Quiz

Some questions to verify that you understand the way we use graphs:

```{r quiz}
quiz(
  question("Which package contains functions for plotting graphs?",
    answer("base"),
    answer("graph"),
    answer("igraph", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the objects listed below are components of a graph?",
    answer("vertex", correct = TRUE),
    answer("border"),
    answer("center"),
    answer("edge", correct = TRUE)
  )
)
```
